---
title: "Lifetime brain atrophy is correlated with longitudinally-observed atrophic changes"
format: 
  html:
    code-fold: true
author: "Anna Elisabeth Furtjes"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html
doi: test
---

```{r setup, include=F, warning=F, message=F, eval=T}
load("paths.RData")
knitr::opts_knit$set(root.dir = wd)
#quarto::quarto_render(input = "long_assocs.qmd", output_file = "long_assocs.html")
```

Results produced by the code below are described in the manuscript under section:\
**LBA moderately  captured observed atrophic changes that were longitudinally observed between two MRI scans nine years apart**.

The outcome traits were derived using code displayed in 'Data preparation': [LBC](test.html) and [UKB](test.html).
Ratio and residual scores were not flipped in these analyses, like they were in the analyses presented in the main manuscript. This does not affect the directions of effects when we were comparing e.g., LBA residual score with longitudinal residual score.

# Load packages

```{r packages, warning=F, message=F, eval=T}
library(ggplot2)
library(data.table)
library(ggpubr)
library(patchwork)
```

# Define function
```{r, warning=F, message=F}
plot_heatmap = function(dat = both[,c("ICV","TBV", "diff", "ratio", "resid")], axisNames = c("ICV","TBV","Difference\nscore","Ratio\nscore","Residual\nscore")){

  # define function to obtain lower triangle
  get_lower_tri<-function(cormat){
    cormat[upper.tri(cormat)] <- NA
    return(cormat)
  }

  # get correlation matrix for both samples together
  cor = cor(dat, use = "pairwise.complete.obs")
  cor = get_lower_tri(cor)
  # melt matrix
  melted = reshape2::melt(cor)
  # get rounded value
  melted$value_round = round(melted$value, digit = 2)
  melted$distance0 = abs(melted$value)

  # plot
  library(ggplot2)

  p = ggplot(data = melted)+
    geom_point(aes(x = Var1, y = Var2, shape = value, fill = value, size = distance0), shape = 21, alpha = 0.7, colour = "white") +
    scale_fill_gradient2(low = "#82A0D8", high = "#8DDFCB", mid = "white",
   midpoint = 0, limit = c(-1,1), space = "Lab" ,name="Correlation", guide = "legend")+
    scale_size_continuous(range = c(1, 15), guide = "none")+
    geom_text(aes(Var1, Var2, label = value_round), color = "black", size = 4)+
    xlab("")+
    ylab("")+
    #ggtitle("Pearson's corelations")+
    scale_x_discrete(labels = axisNames)+
    scale_y_discrete(labels = axisNames)+
    guides(fill = "none")+
    theme_bw()+
    theme(panel.border = element_blank())

  return(p)
}
```


# Pearson's correlations

## LBC

Cross-sectional estimates considered from final neuroimaging visit (4th scan, 5th visit)


```{r, message=F, warning=F, fig.dim=c(6,6)}
# read in LBC data
## longitudinal
long = fread(paste0(out, "/LBC1936_longTBVWaves2and5.txt"), select = c("lbc36no","TBVdiff_2to5","TBVratio_5to2","TBVresid_2to5"))
## cross-sectional
cross4 = fread(paste0(out, "/LBC1936_crossNeuroWave4.txt"), select = c("lbc36no","ICV","TBV", "diff", "ratio", "resid"))

# merge 
LBC = merge(long, cross4, by = "lbc36no")

# number of participants
n = nrow(LBC)

# plot correlations between atrophy measures
p = plot_heatmap(dat = LBC[,c("ICV","TBV", "diff", "ratio", "resid","TBVdiff_2to5","TBVratio_5to2","TBVresid_2to5")], 
                 axisNames = c("ICV","TBV","Difference\nscore\n(cross)","Ratio\nscore\n(cross)","Residual\nscore\n(cross)","Difference\nscore\n(long)","Ratio\nscore\n(long)","Residual\nscore\n(long)"))+
  ggtitle(paste0("LBC Wave 5\n(N = ",n,")"))
p

#ggsave(paste0(out, "phenotypic/LBC_corPlot.jpg"), plot = p, width = 15, height = 15, units = "cm", dpi = 300)
```


Residual score:

  - r = `r round(cor.test(LBC$resid, LBC$TBVresid_2to5)$estimate, digits = 2)`; **p** = `r signif(cor.test(LBC$resid, LBC$TBVresid_2to5)$p.value, digits = 2)`


Ratio score:

  - r = `r round(cor.test(LBC$ratio, LBC$TBVratio_5to2)$estimate, digits = 2)`; **p** = `r signif(cor.test(LBC$ratio, LBC$TBVratio_5to2)$p.value, digits = 2)`


Difference score:

  - r = `r round(cor.test(LBC$diff, LBC$TBVdiff_2to5)$estimate, digits = 2)`; **p** = `r signif(cor.test(LBC$diff, LBC$TBVdiff_2to5)$p.value, digits = 2)`


## UKB

Cross-sectional estimates considered from second neuroimaging visit

```{r corrUKB, message=F, warning=F, fig.dim=c(6,6)}
# read in UKB neuro data
UKB = fread(paste0(out,"/UKB_neuroNoLongProcess.txt"))

# restrict to second neuroimaging visit (i.e., third visit altogether)
UKB3 = UKB[UKB$wave == 3,]

# later edit: exclude extreme outliers from longitudinal data 
UKB3 <- UKB3[which(UKB3$TBVdiff_2to3_stand < 10),]
UKB3 <- UKB3[which(UKB3$TBVdiff_2to3_stand > (-10)),]

UKB3 <- UKB3[which(UKB3$TBVratio_3to2_stand < 10),]
UKB3 <- UKB3[which(UKB3$TBVratio_3to2_stand > (-10)),]

UKB3 <- UKB3[which(UKB3$TBVresid_2to3_stand < 10),]
UKB3 <- UKB3[which(UKB3$TBVresid_2to3_stand > (-10)),]

# now that more participants were excluded, need to re-calculate the residual score
model <- lm(TBV ~ ICV, data = UKB3)
UKB3$resid = resid(model)

UKB3$resid_stand <- as.vector(scale(UKB3$resid))

# number of participants
n = nrow(UKB3)

# plot correlations between atrophy measures
p = plot_heatmap(dat = UKB3[,c("ICV", "TBV", "CSF", "diff", "ratio", "resid", "TBVdiff_2to3", "TBVratio_3to2", "TBVresid_2to3")], 
                 axisNames = c("ICV", "TBV", "CSF", "Difference\nscore\n(cross)", "Ratio\nscore\n(cross)", "Residual\nscore\n(cross)", "Difference\nscore\n(long)", "Ratio\nscore\n(long)", "Residual\nscore\n(long)"))+
  ggtitle(paste0("UKB\nsecond neuroimaging visit\n(N = ",n,")"))
p

#ggsave(paste0(out, "UKB_corPlot.jpg"), plot = p, width = 15, height = 15, units = "cm", dpi = 300)
```

Residual score:

  - r = `r round(cor.test(UKB3$resid, UKB3$TBVresid_2to3)$estimate, digits = 2)`; **p** = `r signif(cor.test(UKB3$resid, UKB3$TBVresid_2to3)$p.value, digits = 2)`


Ratio score:

  - r = `r round(cor.test(UKB3$ratio, UKB3$TBVratio_3to2)$estimate, digits = 2)`; **p** = `r signif(cor.test(UKB3$ratio, UKB3$TBVratio_3to2)$p.value, digits = 2)`


Difference score:

  - r = `r round(cor.test(UKB3$diff, UKB3$TBVdiff_2to3)$estimate, digits = 2)`; **p** = `r signif(cor.test(UKB3$diff, UKB3$TBVdiff_2to3)$p.value, digits = 2)`


### Repeat heatmap for residual score derived with T1-scaling factor

```{r scaling,message=F, warning=F, fig.dim=c(6,6)}
### make same plot only for cross-sectional measures to compare to resid Scaling factor
scaling = plot_heatmap(dat = UKB3[,c("ICV", "TBV", "T1ScalingFactor", "diff", "ratio", "resid", "residScalingFactor_stand")], 
                 axisNames = c("ICV", "TBV", "T1 Scaling Factor", "Difference\nscore\n(cross)", "Ratio\nscore\n(cross)", "Residual\nscore\n(cross)", "Residual score\n(scaling factor)"))+
  ggtitle(paste0("UKB\nsecond neuroimaging visit\n(N = ",n,")"))
scaling

#ggsave(paste0(out, "phenotypic/UKB_scalingFactor.jpg"), plot = scaling, width = 14, height = 14, units = "cm", dpi = 200)
```

