---
title: "Human Connectome Project"
format: 
  html:
    code-fold: true
author: "Anna Elisabeth Furtjes"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html
doi: test
---

------------------------------------------------------------------------

```{r setup, include=F, warning=F, message=F, eval=T}
load("paths.RData")
knitr::opts_knit$set(root.dir = wd)
#quarto::quarto_render(input = "HCP.qmd", output_file = "HCP.html")
```

# Load packages

```{r packages, message=F, warning=F, include = T}
library(data.table)
library(ggplot2)
library(stringr)
library(ggpubr)
library(cowplot)
library(dplyr)
```

# Read in and format the data
```{r, message=F, warning=F}
HCPcov = fread(paste0(HCPdir,list.files(path = HCPdir, pattern = "unrestricted_annafurtjes")))
HCPcov = HCPcov[,c("Subject", "Release", "Acquisition","fMRI_3T_ReconVrs")]

HCP = fread(paste0(HCPdir,"unrestricted_hcp_freesurfer.csv"))
HCP = HCP[,c("Subject", "Gender", "FS_InterCranial_Vol", "FS_BrainSeg_Vol_No_Vent")]

HCP = merge(HCP, HCPcov, by = "Subject")

HCPage = fread(paste0(HCPdir, "RESTRICTED_annafurtjes_12_14_2023_4_18_2.csv"))
HCPage = HCPage[,c("Subject","Age_in_Yrs", "Family_ID")]

HCP = merge(HCP, HCPage, by ="Subject")
names(HCP) = c("ID", "Sex", "ICV", "TBV", "Release", "Acquisition","fMRI_3T_ReconVrs", "Age", "Family_ID")

# convert mm3 estimates to more intuitive cm3 estimates
HCP$ICV = HCP$ICV/1000
HCP$TBV = HCP$TBV/1000

# calculate atrophy scores
HCP$diff = HCP$ICV - HCP$TBV
HCP$ratio = HCP$TBV / HCP$ICV

# delete those from data 
HCP=HCP[-which(HCP$diff < 0),]

# estimate residual model
model <- lm(TBV ~ ICV, data = HCP)
HCP$resid = as.vector(resid(model, na.rm=T))

# standardise variables
HCP$diff_stand = as.vector(scale(HCP$diff))
HCP$ratio_stand = as.vector(scale(HCP$ratio))
HCP$resid_stand = as.vector(scale(HCP$resid))

```



# Age correlations in the full sample

It is contrary to our expectation that ICV gets so much larger with younger age - cohort effect?

  - Much stronger correlation between younger age and ICV than there should be (N = 1103)
  
  - This explains why the atrophy scores with dependence of baseline levels are also strongly affected by age
  
  - The only score that is independent of baseline levels - i.e., residual score - has no association with age


```{r, message=F, warning=F}
corICV = ggplot(data = HCP, aes(x = ICV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())

corTBV = ggplot(data = HCP, aes(x = TBV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())

cordiff = ggplot(data = HCP, aes(x = diff, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Difference score")+
  theme(panel.border = element_blank())

corratio = ggplot(data = HCP, aes(x = ratio, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Ratio score")+
  theme(panel.border = element_blank())

corresid = ggplot(data = HCP, aes(x = resid, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Residual score")+
  theme(panel.border = element_blank())

ggarrange(corICV, corTBV, cordiff, corratio, corresid, common.legend=T, legend = "bottom")
```




# Excluding related individuals

  - Relatedness is determined based on Family_ID variable
  
  - The correlations stay about the same even when excluding all related individuals (N = 444)
  
```{r, message=F, warning=F}
##### Family_ID
## Randomly select one participant from each family and keep  only one
HCPunrel = HCP[, .SD[sample(x = .N, size = 1)], by = Family_ID]

# calculate atrophy scores
HCPunrel$diff = HCPunrel$ICV - HCPunrel$TBV
HCPunrel$ratio = HCPunrel$TBV / HCPunrel$ICV

# delete those from data (none in this unrelated data)
if(sum(HCPunrel$diff < 0) != 0){HCPunrel=HCPunrel[-which(HCPunrel$diff < 0),]}

# estimate residual model
model <- lm(TBV ~ ICV, data = HCPunrel)
HCPunrel$resid = as.vector(resid(model, na.rm=T))

# standardise variables
HCPunrel$diff_stand = as.vector(scale(HCPunrel$diff))
HCPunrel$ratio_stand = as.vector(scale(HCPunrel$ratio))
HCPunrel$resid_stand = as.vector(scale(HCPunrel$resid))


########## plot age correlations (copied from Aim 1)
corICV = ggplot(data = HCPunrel, aes(x = ICV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())

corTBV = ggplot(data = HCPunrel, aes(x = TBV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())


cordiff = ggplot(data = HCPunrel, aes(x = diff, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Difference score")+
  theme(panel.border = element_blank())

corratio = ggplot(data = HCPunrel, aes(x = ratio, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Ratio score")+
  theme(panel.border = element_blank())

corresid = ggplot(data = HCPunrel, aes(x = resid, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Residual score")+
  theme(panel.border = element_blank())

ggarrange(corICV, corTBV, cordiff, corratio, corresid, common.legend=T, legend = "bottom")
```



# Adjusting for batch effects

Adjusting for batch effects does not change the strong age correlation we see in this young sample either.


```{r, message=F, warning=F}
# adjust for Release and acquisition

# adjust ICV
model = lm('ICV ~ Release + Acquisition + fMRI_3T_ReconVrs', data =HCP)
summary(model)

HCP$ICVadj = resid(model)

corICV = ggplot(data = HCP, aes(x = ICVadj, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("ICV adjusted for Release and Acquisition")+
  theme(panel.border = element_blank())

# adjust TBV 
model = lm('TBV ~ Release + Acquisition', data =HCP)
HCP$TBVadj = resid(model)

corTBV = ggplot(data = HCP, aes(x = TBVadj, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("TBV adjusted for Release and Acquisition")+
  theme(panel.border = element_blank())

# re-create diff score with adjusted variables
HCP$diff_adj = HCP$ICVadj - HCP$TBVadj
  
cordiff = ggplot(data = HCP, aes(x = diff_adj, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Difference score")+
  theme(panel.border = element_blank())

# re-create ratio score with adjusted variables
HCP$ratio_adj = HCP$TBVadj / HCP$ICVadj
 
corratio = ggplot(data = HCP, aes(x = ratio_adj, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Ratio score")+
  theme(panel.border = element_blank())

ggarrange(corICV, corTBV, common.legend = T, legend = "bottom")
```



# Restrict sample by age

  - If we consider participants of and below 29 years only, the ICV age correlation disappears
  
  - This threshold was empirically determined and then applied in the plot below 
  
  - In this younger sample, we see no age correlations whatsoever anymore
  

## Determine most appropriate age cut-off for HCP

  - the plot shows that the ICV-age correlation disappears at age 29 and below
  
  - also this sample includes only 3 out of the 10 participants that had to get excluded due to ICV < TBV
  

```{r, message=F, warning=F}
### determne which age is a good one to use for cut-off
# determine age cut-offs to iterate through
ageCut = seq(from = min(HCP$Age)+1, to = max(HCP$Age), by = 1)

# object to store results
storeNames = c("Age cut-off value", "Cor", "p", "ci_l", "ci_u", "Measure")
store = as.data.frame(matrix(nrow = length(ageCut), ncol = length(storeNames)))
names(store) = storeNames

# iterate over each age cut-off and calculate scores
for(i in ageCut){
  # store which age cut off iteration this is
  loc = which(is.na(store$`Age cut-off value`))[1]
  store[loc,"Age cut-off value"] = i
  
  # cut sample
  Youngdata = HCP[which(HCP$Age <= i),]
  
  # calculate correlations
  ## ICVerence
  store[loc,"Cor"] =
    with(Youngdata, cor.test(Age, ICV))$estimate
  
  store[loc,"p"] =
    with(Youngdata, cor.test(Age, ICV))$p.value
  
  store[loc,"ci_l"] =
    with(Youngdata, cor.test(Age, ICV))$conf.int[1]
  
  store[loc,"ci_u"] =
    with(Youngdata, cor.test(Age, ICV))$conf.int[2]
  
  store[loc,"Measure"] = "ICV"
}

ggplot(data = store)+
  geom_point(aes(x = Cor, y = `Age cut-off value`), alpha = 0.5)+
  geom_errorbar(aes(y = `Age cut-off value`, xmin = ci_l, xmax = ci_u), alpha = 0.3)+
  geom_vline(xintercept = -0.1, color = "grey")+
  geom_hline(yintercept = 31, color = "grey")+
  xlab("Correlation with age")+
  ylab("Age cut-off used to subset\nsample into a younger subsample")+
  scale_y_continuous(limits = c(21, 37.5), breaks = seq(from = 20, to = 36, by = 2))+
  ggtitle("Age-ICV correlation in HCP")+
  theme_bw()+
  theme(panel.border = element_blank())
```



## Explore age correlations in sample of or below 31 years

  - When we restrict our sample to <= 31 years, the age correlations disappear

```{r, message=F, warning=F}
HCPyoung = HCP[which(HCP$Age <= 31),]

# calculate atrophy scores
HCPyoung$diff = HCPyoung$ICV - HCPyoung$TBV
HCPyoung$ratio = HCPyoung$TBV / HCPyoung$ICV

# delete those from data 
#HCPyoung=HCPyoung[-which(HCPyoung$diff < 0),]

# estimate residual model
model <- lm(TBV ~ ICV, data = HCPyoung)
HCPyoung$resid = as.vector(resid(model, na.rm=T))


corICV = ggplot(data = HCPyoung, aes(x = ICV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())

corTBV = ggplot(data = HCPyoung, aes(x = TBV, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  theme(panel.border = element_blank())


cordiff = ggplot(data = HCPyoung, aes(x = diff, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Difference score")+
  theme(panel.border = element_blank())

corratio = ggplot(data = HCPyoung, aes(x = ratio, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Ratio score")+
  theme(panel.border = element_blank())

corresid = ggplot(data = HCPyoung, aes(x = resid, y = Age))+
  geom_point(alpha = 0.2)+
  theme_bw()+
  geom_smooth(method='lm')+
  stat_cor(method = "pearson",color ="blue")+
  ylab("Age in years")+
  xlab("Residual score")+
  theme(panel.border = element_blank())

ggarrange(corICV, corTBV, cordiff, corratio, corresid, common.legend=T, legend = "bottom")

```

# Plot distributions colored by age

Shown in Supplementary Figure 3: Distributions of TBV, ICV, and lifetime brain atrophy estimated with the residual, ratio, and difference method. Histograms are coloured by age groups. 


```{r, eval = T, message=F, warning=F,fig.dim=c(20,8)}
####################################################
# read in HCP data
HCP = fread(paste0(HCPdir,"/unrestricted_hcp_freesurfer.csv"))
HCP = HCP[,c("Subject", "Gender", "FS_InterCranial_Vol", "FS_BrainSeg_Vol_No_Vent")]
names(HCP) = c("ID", "Sex", "ICV", "TBV")

# add age information
HCPage = fread(paste0(wd, "/RESTRICTED_annafurtjes_12_14_2023_4_18_2.csv"))
names(HCPage)[which(names(HCPage) == "Subject")] = "ID"
names(HCPage)[which(names(HCPage) == "Age_in_Yrs")] = "Age"
HCP = merge(HCP, HCPage[,c("ID","Age")], by = "ID")

# as outlined elsewhere, empirical investigations warrant to use an age cut-off of 31 years in this sample
HCP = HCP[which(HCP$Age <= 31),]

# convert mm3 estimates to more intuitive cm3 estimates
HCP$ICV = HCP$ICV/1000
HCP$TBV = HCP$TBV/1000

# estimate brain atrophy from single MRI scan
HCP$diff = HCP$ICV - HCP$TBV
HCP$ratio = HCP$TBV / HCP$ICV

# Quality control: 
#print(paste0("Some participants have negative difference scores and ratio scores > 1, which means that their ICV estimate is smaller than their TBV estimate. This must be an error as the skull always surrounds the brain. Those ", sum((HCP$diff < 0))," HCP participants were excluded from the data set."))

deletedHCP = sum(HCP$diff < 0)
# delete those from data 
if(sum(HCP$diff < 0) != 0){HCP=HCP[-which(HCP$diff < 0),]}

# estimate residual model
model <- lm(TBV ~ ICV, data = HCP)
HCP$resid = as.vector(resid(model, na.rm=T))

  
# standardise variables
HCP$diff_stand = as.vector(scale(HCP$diff))
HCP$ratio_stand = as.vector(scale(HCP$ratio))
HCP$resid_stand = as.vector(scale(HCP$resid))

####################################################
# make age groups
HCP$Age_group <- NA
HCP$Age_group[HCP$Age < 24] <- "23 years and under"
HCP$Age_group[HCP$Age >= 24 & HCP$Age < 25] <- "24"
HCP$Age_group[HCP$Age >= 25 & HCP$Age < 26] <- "25"
HCP$Age_group[HCP$Age >= 26 & HCP$Age < 27] <- "26"
HCP$Age_group[HCP$Age >= 27 & HCP$Age < 28] <- "27"
HCP$Age_group[HCP$Age >= 28 & HCP$Age < 29] <- "28"
HCP$Age_group[HCP$Age >= 29] <- "29 years and over"

p1=ggplot(HCP, aes(x=TBV, fill=Age_group)) +
  geom_histogram()+
  scale_fill_manual("Age groups", values = c("#292f56", "#1e4572", "#005c8b", "#008ba0", "#00bca1","#69e882", "#acfa70"))+
  xlab("TBV")+
  theme_bw()

p2=ggplot(HCP, aes(x=ICV, fill=Age_group)) +
  geom_histogram()+
  scale_fill_manual("Age groups", values = c("#292f56", "#1e4572", "#005c8b", "#008ba0", "#00bca1","#69e882", "#acfa70"))+
  xlab("ICV")+
  theme_bw()

p3=ggplot(HCP, aes(x=resid_stand, fill=Age_group)) +
  geom_histogram()+
  scale_fill_manual("Age groups", values = c("#292f56", "#1e4572", "#005c8b", "#008ba0", "#00bca1","#69e882", "#acfa70"))+
  xlab("Residual score")+
  theme_bw()

p4=ggplot(HCP, aes(x=ratio_stand, fill=Age_group)) +
  geom_histogram()+
  scale_fill_manual("Age groups", values = c("#292f56", "#1e4572", "#005c8b", "#008ba0", "#00bca1","#69e882", "#acfa70"))+
  xlab("Ratio score")+
  theme_bw()

p5=ggplot(HCP, aes(x=diff_stand, fill=Age_group)) +
  geom_histogram()+
  scale_fill_manual("Age groups", values = c("#292f56", "#1e4572", "#005c8b", "#008ba0", "#00bca1","#69e882", "#acfa70"))+
  xlab("Difference score")+
  theme_bw()

pHCP <- ggarrange(p1,p2,p3,p4,p5, nrow = 1, common.legend = T, legend = "bottom")
# add title
pHCP <- annotate_figure(pHCP, top = text_grob("HCP",face = "bold", size = 14))

#ggsave(paste0(out,"phenotypic/HCP_disttributions.jpg"), bg = "white",plot = pHCP, width = 30, height = 10, units = "cm", dpi = 300)
pHCP
```
